---
title: "Use biomaRt with Rattus norvegicus"
date: "`r Sys.Date()`"
execute:
  echo: true
  cache: false
format:
  html:
    mermaid:
      theme: neutral
    toc: true
    embed-resources: true
    number-sections: true
    html-math-method: katex
    highlight-style: github
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

To begin, install the {biomaRt} package.

```{r install_biomart, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biomaRt")
```

## Package

Load package.

```{r load_biomart}
packageVersion("biomaRt")
suppressPackageStartupMessages({
  library(biomaRt)
})
```

## Ensembl 114

List archives.

```{r list_ensembl_archives}
biomaRt::listEnsemblArchives()
```

Use Ensembl 114.

```{r ensembl_114}
my_host <- "https://may2025.archive.ensembl.org"
ensembl_114 <- useMart("ENSEMBL_MART_ENSEMBL", host = my_host)
avail_datasets_v114 <- listDatasets(ensembl_114)
rat_dataset <- grep('norvegicus', avail_datasets_v114$dataset, value = TRUE)
rat_dataset
```

Rat database.

```{r ensembl_114_rat}
ensembl_114_rat <- useMart(
  biomart = "ENSEMBL_MART_ENSEMBL",
  dataset = rat_dataset,
  host = my_host
)
```

## Database results

Load database results.

```{r manual_res}
manual_res <- readr::read_tsv(file = "../rattus_norvegicus_core_114_1_go_lookup.txt", show_col_types = FALSE)
head(manual_res)
```

The manual results are not unique!

```{r manual_res_unique}
manual_res |>
  dplyr::arrange(ensembl_gene_id, go_id) |>
  dplyr::add_count(across(everything())) |>
  head(10)
```

Create unique table.

```{r manual_res_uniq}
manual_res |>
  dplyr::arrange(ensembl_gene_id, go_id) |>
  dplyr::distinct() -> manual_res_uniq

head(manual_res_uniq)
```

Number of genes.

```{r manual_res_uniq_genes}
length(unique(manual_res_uniq$ensembl_gene_id))
```

## Gene Ontology lookup

Get all Ensembl Gene IDs.

```{r all_ensembl_gene_id}
all_gene_ids <- getBM(
  attributes = "ensembl_gene_id",
  mart = ensembl_114_rat
)

length(all_gene_ids$ensembl_gene_id)
length(unique(all_gene_ids$ensembl_gene_id))
```

Using `ensembl_gene_id` get all `go_id`.

```{r ensembl_gene_id_and_go_id}
gene_id_go <- getBM(
  attributes = c("ensembl_gene_id", "go_id"),
  filters = "ensembl_gene_id",
  values = all_gene_ids$ensembl_gene_id,
  mart = ensembl_114_rat
)

dim(gene_id_go)
head(gene_id_go)
```

Number of genes.

```{r gene_id_go_uniq_genes}
length(unique(gene_id_go$ensembl_gene_id))
```

Need to remove empty entries.

```{r remove_empty_go_id}
gene_id_go |>
  dplyr::filter(go_id != "") -> gene_id_go_clean

length(unique(gene_id_go_clean$ensembl_gene_id))
```

## Compare Gene Ontology tables

Each row in both `gene_id_go` and `manual_res` represents a unique `(ensembl_gene_id, go_id)` pair. To compare the two tables we use two `dplyr` verbs:

- `dplyr::intersect(x, y)` treats the rows of each table as a set and returns only those rows that appear in **both** `x` and `y`. Both columns must match for a row to be included.
- `dplyr::anti_join(x, y, by)` returns all rows from `x` that have **no match** in `y` based on the specified key columns. It is applied twice: once to find pairs present only in `gene_id_go` (biomaRt result) and once to find pairs present only in `manual_res` (database result).

Together, these three operations partition all observed pairs into three disjoint groups: shared, biomaRt-only, and manual-only.

```{r compare_gene_ontology_tables}
both      <- dplyr::intersect(gene_id_go_clean, manual_res_uniq)
only_bm   <- dplyr::anti_join(gene_id_go_clean, manual_res_uniq, by = c("ensembl_gene_id", "go_id"))
only_man  <- dplyr::anti_join(manual_res_uniq, gene_id_go_clean, by = c("ensembl_gene_id", "go_id"))

cat("Pairs in both:         ", nrow(both), "\n")
cat("Only in biomaRt:       ", nrow(only_bm), "\n")
cat("Only in manual result: ", nrow(only_man), "\n")
```

```
Pairs in both:          210063 
Only in biomaRt:        0 
Only in manual result:  0 
```
